% ======================= CHAPTER 5: CONCLUSION ==========================
\chapter{Conclusion and Future Work}

\section{Summary of Contributions}

This thesis trained and evaluated four deep learning architectures---EfficientNet-B0, ResNet50, DenseNet121, and Swin-Tiny---on the ISIC 2018 seven-class skin lesion dataset, with a training pipeline built specifically around the dataset's 58:1 class imbalance. The pipeline's four components each address the imbalance from a different angle: Weighted Random Sampling at the data-loading level, Mixup and CutMix at the training-signal level, label smoothing at the target level, and cosine-annealed AdamW with weight decay at the optimisation level. Together they eliminate the majority-class collapse that appears in naive training, as confirmed by the ablation study.

At the single representative seed (42) reported in the headline tables, EfficientNet-B0 (5.3M parameters) outperformed ResNet50 (25.6M), DenseNet121 (8M), and Swin-Tiny (28M) on every hard-classification metric. When the experiment was repeated across five random seeds, however, Swin-Tiny emerged as the strongest individual backbone on average ($88.33 \pm 1.96\%$ accuracy versus EfficientNet-B0's $87.31 \pm 1.32\%$), with the seed-42 ranking reversing because of Swin-Tiny's higher variance. The lesson is the same: when training data is limited, architectural efficiency and training methodology matter more than raw parameter count, but a single-seed comparison is not enough to rank closely-matched architectures. The soft-voting ensemble showed that the four architectures are complementary rather than redundant---combining them raised macro AUC from 0.952 to 0.980 and melanoma F1-score from 0.695 to 0.744 at the headline seed, and held a $89.93 \pm 0.66\%$ accuracy across the five seeds, with McNemar's test confirming the gain over every single model is statistically significant ($p < 10^{-7}$ against the strongest baseline). The improvement comes from architectural diversity rather than additional training cost.

Grad-CAM visualisations confirmed that every model attends to the lesion interior and boundary, not to background skin or imaging artefacts. This is a necessary condition for clinical deployment---a model that achieves 88\% accuracy by attending to ruler markings is not useful, regardless of its test-set numbers. An interactive Streamlit prototype embedding real-time ensemble prediction and Grad-CAM overlay demonstrates that these models can be integrated into a functional diagnostic support interface.

Beyond the image-only baseline, two further experiments were conducted to extend and probe the result. Late-fusing the HAM10000 patient metadata vector (age, sex, anatomical location) into each backbone through a small MLP encoder pushes the four-model ensemble to 90.49\% accuracy and 0.873 macro F1, improving on Pacheco and Krohling's reported 89.0\% / 0.83 and crossing the 90\% threshold often cited as a reference point for image-based skin classification. To bound expectations honestly, the image-only ensemble was also evaluated zero-shot on PAD-UFES-20, a public dataset of 2,298 clinical smartphone photographs from a different acquisition modality. Accuracy drops to 32.5\%, a 56-percentage-point fall that is consistent with literature reports for dermoscopy-to-clinical-photo transfer without domain adaptation. Per-class AUC remains in the 0.61--0.81 range, confirming the model is not random on smartphone photographs but that hard-decision performance requires modality-specific adaptation or fine-tuning to be clinically usable.

\section{Key Findings}

Three findings from this work have implications beyond the specific dataset and architectures studied.

The accuracy--AUC inversion across models is the most practically relevant. CNNs (EfficientNet-B0 in particular) produce sharper hard decisions while Swin-Tiny produces better-calibrated probability distributions, and this difference is not just a matter of scale---it persists after identical training on identical data. The implication for system design is that the appropriate model depends on what the downstream clinical task actually requires. A binary flag for immediate referral calls for EfficientNet-B0's high F1. A ranked worklist for a screening programme calls for Swin-Tiny's superior AUC. Choosing based on either metric alone, without understanding the deployment context, risks optimising the wrong objective.

The label smoothing result from the ablation study is counterintuitive enough to warrant separate attention. Removing label smoothing simultaneously increased AUC (from 0.952 to 0.967) and decreased macro F1 (by 3.1 percentage points). Label smoothing trades calibration sharpness for decision boundary quality---it prevents confident predictions that would raise AUC but helps the model generalise the ambiguous boundaries between MEL, NV, and BKL. The conventional advice to apply label smoothing universally is therefore too coarse; the right choice depends on whether the model's output will be thresholded or used as a continuous score.

Melanoma remains the hardest class after all interventions: the ensemble's MEL F1-score of 0.744 is the lowest among all well-represented classes, even after combining four architectures with complementary inductive biases. This is not a failure of the training pipeline. It reflects the genuine clinical difficulty of separating early-stage melanoma from benign nevi---a task that experienced dermatologists also find unreliable---and suggests that further progress on this specific class requires either melanoma-targeted feature learning, segmentation supervision for lesion boundaries, or substantially more MEL training data.

\section{Limitations}

Several constraints limit the conclusions that can be drawn from this study:

\begin{itemize}
    \item \textbf{Dataset demographic bias.} HAM10000 was collected primarily in Austria and the United States and reflects a population that is predominantly Fitzpatrick I--III. Deployment in patient cohorts with darker skin tones (Fitzpatrick IV--VI) is likely to underperform, consistent with the skin-tone bias reported in commercial dermoscopy systems. This thesis does not include Vietnamese or otherwise locally representative data, so the demographic gap is acknowledged but not measured.
    \item \textbf{Acquisition modality gap.} The out-of-distribution evaluation on PAD-UFES-20 quantifies a 56-percentage-point accuracy drop when the same image-only ensemble is run on clinical smartphone photographs instead of polarised-light dermoscopy. The pipeline is not deployable on consumer-grade imagery without domain adaptation, in-domain fine-tuning, or modality-specific preprocessing.
    \item \textbf{Closed-set seven-class scope.} The classifier is forced to assign every input to one of seven ISIC categories. Real clinical practice covers dozens of lesion types, and any input outside the trained classes (acne, scars, non-lesion regions) is misclassified into the nearest match. The Streamlit prototype mitigates this with a confidence-threshold warning, but a dedicated open-set or OOD detector is not implemented.
    \item \textbf{Fixed $224 \times 224$ resolution.} Higher resolutions, which architectures such as EfficientNetV2~\cite{tan2021efficientnetv2} are designed to exploit, may recover discriminative features lost at the standard size, particularly for subtle pigmentation patterns and texture details in the MEL/NV/BKL confusion cluster.
    \item \textbf{Small rare-class test sets.} The test set contains 17 Dermatofibroma and 22 Vascular Lesion samples. A single additional misclassification shifts these per-class F1-scores by approximately 0.06, so figures for the rare classes carry high variance and should not be over-interpreted.
    \item \textbf{Single Transformer architecture.} Only Swin-Tiny was evaluated. Other variants (DeiT~\cite{touvron2021deit}, ConvNeXt~\cite{liu2022convnext}, or larger Swin variants) may produce different accuracy--AUC trade-offs and different ensemble characteristics when combined with the CNN backbones.
    \item \textbf{No prospective clinical validation.} The models were not deployed in a patient-facing workflow or evaluated by practising dermatologists in real time. Comparative human--AI studies and clinician acceptance assessments remain prerequisites for any deployment consideration.
\end{itemize}

\section{Future Work}

The most direct extensions of this work address its main limitations:

\begin{itemize}
    \item \textbf{Domain adaptation for clinical photographs.} The 56-percentage-point OOD drop on PAD-UFES-20 establishes a quantitative target. Style-transfer preprocessing, adversarial domain adaptation, or fine-tuning on a small labelled smartphone collection would each be expected to recover a substantial fraction of that gap.
    \item \textbf{Open-set and OOD detection.} Adding a dedicated component (Mahalanobis distance over backbone features, Energy-based score, or autoencoder reconstruction signal) would allow the system to reject inputs that are not skin lesions, rather than forcing a seven-class assignment that the user is then asked to discount.
    \item \textbf{Skin-tone-aware augmentation and fairness analysis.} Fitzpatrick-aware hue and saturation augmentation, combined with subgroup fairness loss terms, would directly target the demographic bias inherited from the HAM10000 training distribution. Once Vietnamese or otherwise locally representative data is available, the resulting fairness numbers can be measured.
    \item \textbf{Higher-resolution training.} Architectures designed for larger inputs may recover discriminative features lost at $224 \times 224$, particularly for the MEL/NV/BKL confusion cluster where small texture differences carry diagnostic information.
    \item \textbf{Self-supervised pretraining.} Methods such as DINO~\cite{caron2021dino} can extract strong visual representations from large unlabelled dermoscopy collections before fine-tuning on labelled data, potentially reducing the effective data requirement and improving minority-class generalisation.
    \item \textbf{Segmentation-assisted classification.} Jointly predicting lesion boundaries~\cite{hasan2020dsnet} alongside the disease label would give the model explicit spatial supervision for the boundary features most relevant to the melanoma--nevus distinction.
    \item \textbf{Knowledge distillation for deployment.} Compressing the four-model ensemble into a single lightweight model would preserve most of the accuracy gain while reducing inference cost to a single forward pass, a practical requirement for mobile or point-of-care deployment.
    \item \textbf{Prospective clinical evaluation.} A controlled study comparing the system's outputs against a panel of practising dermatologists on a representative case set is the necessary next step before any deployment claim can be made.
\end{itemize}
